---
title: "Data wrangling II"
author: ""
date: ""
output: html_document
editor_options: 
  chunk_output_type: console
---

## Main ideas

- To answer questions with data, we often need to use related data from many 
  different datasets.
  
- We can combine data from different sources using a well-chosen join function.

## Packages

We'll make use of the following packages today.

```{r packages, message=FALSE}
library(tidyverse)
#install.packages(nycflights13)
library(nycflights13)  
```

## Lecture notes and exercises

Common `dplyr` single data frame functions / verbs:

| Function                       | Description                          |
|--------------------------------|--------------------------------------|
| `filter()`                     | pick rows matching criteria          |
| `slice()`                      | pick rows using indices              |
| `select()`                     | pick columns by name                 |
| `pull()`                       | grab a column as a vector            |
| `rename()`                     | rename specific columns              |
| `arrange()`                    | reorder rows                         |
| `mutate()`                     | add new variables                    |
| `transmute()`                  | create new data frame with variables |
| `distinct()`                   | filter for unique rows               |
| `sample_n()` / `sample_frac()` | randomly sample rows                 |
| `summarise()`                  | reduce variables to values           |

Instead of working with a single dataset, usually you will have to work with 
many different related datasets. To answer research questions using related 
datasets, we need to develop tools to join datasets together.

There are many possible types of joins. All have the format 
`something_join(x, y)`.

|  Join function | Description                                                  |
|----------------|--------------------------------------------------------------|
| `inner_join()` | join all rows from `x` where there are matching values in `y`|
|  `left_join()` | include all rows from `x`                                    |
| `right_join()` | include all rows from `y`                                    |
|  `full_join()` | include all rows in `x` or `y`                               |
|  `semi_join()` | return all rows from `x` with match in `y`                   |
|  `anti_join()` | return all rows from `x` without a match in `y`              |

*Note:* `inner_join()` returns all combinations in cases of multiple matches      

```{r test-joins}
x <- tibble(value = c(1, 2, 3),
            xcol  = c("x1", "x2", "x3"))

y <- tibble(value = c(1, 2, 4),
            ycol  = c("y1", "y2", "y4"))

x
y
```

We will demonstrate each of the joins on these small, toy datasets. Check out
the slides for an animated version of these joins.

```{r inner-join}
x
y

inner_join(x, y)
```

```{r left-join}
x
y

left_join(x, y)
```

```{r right-join}
x
y

right_join(x, y)
```

```{r full-join}
x
y

full_join(x, y)
```

```{r semi-join}
x
y

semi_join(x, y)
```

```{r anti-join}
x
y

anti_join(x, y)
```

How do the join functions above know to join `x` and `y` by `value`? Examine the
names to find out.

```{r names}
names(x)
names(y)
```

We will again work with data from the `nycflights13` package.

Examine the documentation for the datasets `airports`, `flights`, and `planes`.

**Question:** How are these datasets related? Suppose you wanted to make a map
of the route of every flight. What variables would you need from which 
datasets?

```{r}
names(flights)
names(airports)
names(planes)
```

- `flights` and `planes` are linked by `tailnum`
- `flights` and `airports` are linked through the FAA codes - `dest` in `flights`
  and `faa` in airports

Left join flights and airports. Note these two datasets have no variables in common
so we will have to specify the variable to join by using `by = `. Check out
the documentation for more information.

```{r flights-join}
flights %>% 
  left_join(airports, by = c("dest" = "faa"))
```

### Practice

(1) Create a new tibble named `dest_delays` with the median arrival delay for 
    each destination. Note this question does not require you to use joins.

```{r make-dest-delays}
dest_delays <- flights %>%
  group_by(dest) %>%
  summarise(delay = median(arr_delay, na.rm = TRUE))
```

(2) Create a new dataset by joining `dest_delays` and `airports`. Only include 
    observations that have both delay and airport information. Note `dest_delays`
    and `airports` have no variables in common so you will need to specify the 
    variables to join using `by` as in the example above.

```{r}
dest_delays %>%
  inner_join(airports, by = c("dest" = "faa")) %>% 
  filter(!is.na(delay))
```

**Question:** Are all of the variables in `dest_delays` included in the new 
dataset you created by joining `dest_delays` and `airports`? Use an appropriate
join function to investigate this issue and determine what is going on here.

There are 105 airports in the `dest_delays` dataset, but when we join to 
airports there are only 100 airports. There are four airports in the 
`dest_delays` data that are not matched in airports. Something weird is going
on.

We use an `anti_join` to help diagnose this issue. Recall `anti_join` returns
all rows from `x` without a match in `y`, so it will return all rows in 
`dest_delays` that don't have a match in `airports`.

```{r}
dest_delays %>%
  anti_join(airports, by = c("dest" = "faa"))
```

A bit of googling reveals that these are airports in Puerto Rico (BQN, PSE, 
SJU) and the U.S. Virgin Islands (STT).

(3) Is there a relationship between the age of a plane and its delays? The plane
    tail number is given in the `tailnum` variable in the `flights` dataset. The 
    year the plane was manufactured is given in the `year` variable in the 
    `planes` dataset. 

- Step #1: Start by finding the average arrival delay for each plane and store
  the resulting dataset in `plane_delays`.

```{r step-1}
plane_delays <- flights %>%
  group_by(tailnum) %>%
  summarise(delay = mean(arr_delay, na.rm = TRUE))
```

- Step #2: Join `plane_delays` to the `planes` data using an appropriate join
  and then use mutate to create an age variable. Note this data is from 2013.

```{r step-2}
plane_delays %>%
  left_join(planes, by = "tailnum") %>%
  mutate(age = 2013 - year)
```

- Step #3: Finally, create an effective visualization of the data.

```{r step-3}
plane_delays %>%
  left_join(planes, by = "tailnum") %>%
  mutate(age = 2013 - year) %>%
  ggplot(aes(x = age, y = delay)) + 
  geom_point(color = "grey30", alpha = 0.5) + 
  geom_smooth(se = FALSE, color = "purple") +
  theme_minimal(base_size = 16) +
  labs(x = "Age", y = "Delay")
```

## Additional Resources

- https://rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf
